???A quantitative framework for investigating the reliability of network construction

### 2017-11-22 manuscript/CI_interactions.tex
added a few more refs

    The assumption of static communities conflicts with known variation in both  the nodes and links included in a network. Community composition and species abundances vary from site to site~\citep{Baiser2012} and over time within a site~\citep{Olesen2011a}. Likewise, interactions over space~\citep{Kitching1987,Baiser2012}, time~\citep{Kitching1987,Olesen2011a}, and between individuals of a given species~\citep{Pires2011a,Fodrie2015,Novak2015}. Moreover, variability in community composition and interactions may or may not be closely related. The removal of a species from a site will obviously also remove its interactions, but the co-occurrence of potentially interacting species do not guarantee they will interact at a given place and time. Interactions can be lost, or not detected, if participating species remain present but do not co-occur temporally, are too rare to detect each other~\citep{Tylianakis2010}, because of environmental contingencies \citep{Poisot2015a}, or through changes to individual preferences~\citep{Fodrie2015}.

    Beyond ``true" variation in network structure, several researchers have pointed to the importance of sampling intensity for the assessment of network structure (e.g.,~\citealp{Martinez1999,Bluthgen2006,XX}). Accumulation of interactions with increasing sampling effort suggests that it is even more challenging to document interactions than species richness \citep{Poisot2012}. As a result, it was proposed that interactions should be described probabilistically and network metrics computed accordingly \citep{Poisot2015b}. Despite this, the acknowledgment that insufficient sampling effort limits our ability to describe variation in ecological networks has not yet led to a quantitative methodology to deal with the uncertainty generated by spatiotemporal variation in ecological interactions and by sampling. Moreover, even in extremely well-sampled networks uneven sampling across species (or pairs of species) can lead to the assumption that some species do not interact because they co-occur rarely or have not yet been observed co-occuring (see Box 1). It is therefore likely that nearly all network studies are missing some interactions and could benefit from an approach that acknowledges this uncertainty.

@@

  We have lots and lots of data, but still not enough for almost all pairs of species.   
% DG : I wonder if a paragraph summarizing the extent of the problem with the salix data should not be provided here. Either within the text, or a box. But we need to convingly show how large is this problem even for such an extensive dataset
% AC : I like the box idea. Since we refer to the Salix data throughout the manuscript it makes sense to make it stand out.

    As a case study, we use an extensively sampled \emph{Salix}-galler-parasitoid meta-network. This dataset consists of a single community type sampled across Europe; willow (\emph{Salix}) species, willow-galling sawflies, and their natural enemies. The data were collected over 29 years at 374 unique locations across Europe, with a total of 641 site visits. Each site visit can be considered as a network in its own right and an independent sample from which to build the meta-network. The meta-network consists of 1,173 different interactions between 52 \emph{Salix} nodes, 92 herbivore nodes, and 126 natural enemy nodes. The high spatiotemporal resolution of this network and the unusually high sampling effort implemented at the site level makes this dataset particularly well suited for illustrating the difficulties in completely sampling a network and testing Bayesian approaches to overcome these difficulties.

    We may begin by comparing the frequency of of co-occurence of taxa to reveal the challenge of having sufficient sampling to be confident that an interaction does not occur (Fig.~\ref{upper_limits}). Most pairs of species (3986/4992 \emph{Salix}-galler pairs and  9794/12096 galler-natural enemy pairs) are never found co-occurring, and for species that did occur the number of co-occurrences was generally low (mean=4.24, var=36.3 for \emph{Salix}-galler pairs; mean=3.87, var=28.8 for galler-natural enemy pairs; Fig.~\ref{histograms}A-B). For those species which interacted when co-occurring, the incidence of interaction was also low (mean=12.0, var=155 for \emph{Salix}-galler pairs; mean=4.04, var=29.3 for galler-natural enemy pairs; Fig.~\ref{histograms}C-D). Only 2.82\% of \emph{Salix}-galler pairs and 7.76\% of galler-natural enemy pairs were observed interacting at all. Thus, even in the most extensive data set that we could find, limited sampling is clearly a major source of uncertainty.

@@

    The combination of these three sources of uncertainty together leads to many potential explanations for the observation of an absence of interaction, but only the situation where $L = 0$ (a true absence) is relevant for the investigation of networks. While the observation of an interaction is straightforward to interpret, absence of interactions must be decomposed into different quantities. It is particularly important to rule out the situations where $D=0 \cup X = 1 \cup L=1$, i.e. the interaction occurred at the location but was not observed, and $D=1 \cup X = 0 \cup L =1$, the interaction is feasible and would have been detected but did not occur at the local site. Ecologists seek to measure the occurrence of true absences, which is the joint event $L=0 \cup X=1 \cup D=1$), but in reality they measure the marginal probability $P(L) = k/n$ where $k$ is again the number of observed interactions and $n$ the number of observed co-occurrences. % I'm pretty sure this should be L=0 but X=1 and D=1: interaction not feasible but site and detection perfect.
    %DG: YES

@@

  An added complication is that not all sources of uncertainty are proportional to sample size. To record an interaction between A and B, we need to identify both partners correctly (a non-trivial problem in many food webs; e.g.~\citet{Kaartinen2011,Roslin2016}) and be able to resolve all interactions with a similar likelihood. For both molecular~\citep{} and rearing~\citep{} techniques, certain types of interactions may go unnoticed due to technical challenges. This can bias the set of recorded interactions.

% DG : I propose to maintain part of this paragraph that was commented out, just to keep the transition. 
The bottom line is that separating different sources of uncertainty is difficult indeed. As an alternative to abandoning empirical networks or continuing to ignore the uncertainty inherent in undetected observations, we propose that some insight regarding the detectability of interactions between species not found co-occurring may be gained from data on other species between which interactions are known to be feasible.

@@

          [[Should we still have this? At the workshop we talked about how interaction probabilities depend on the degrees of both species and so I'm using predicted interaction probabilities (degreeA*degreeB) in the galler example. A writeup for that process follows the original text:]]

@@

          [Tomas adds: This is a bit of a stump]
%DG: I tried to expand just a little, but I don't want to get lost in this paragraph as it would be a distraction. We should basically mention it as a possibility.

@@

    It is fairly straightforward to compute most of network metrics when the different $\lambda$ of the adjacency matrix are known, without variance~\citep{Poisot2016}. Several of these metrics derive directly from quantitative indices of network structure which are equivalent to $\lambda$; the remainder, originally defined for binary networks, can be adjusted to account for interaction probabilities between zero and one. It is not as easy, however, to understand how the uncertainty in these estimated interaction probabilities influences network metrics. Computation of these metrics involves non-linear functions and therefore, because of Jensen's inequality (which states that the average of non-linear function of a stochastic variable differs from the function of the average of that variable), any uncertainty in $\lambda$s could bias both the mean and variance of a network metric. For instance, assume that among all pairs of species with few observed co-occurrences there are many pairs which truly interact. If we begin with no observed links for these pairs, then correcting for these false absences will inflate the number of links and consequently influence metrics such as connectance, nestedness, and degree distribution. While we can, as described above, calculate these network metrics using fixed estimates for $\lambda$, no analytical solution currently exists for uncertain $\lambda$. Instead we must rely on computer simulations. In the following section, we will provide an empirical example, based on the data set published by~\citet{Kopelke2017}. We will show how the uncertainty arising from both absences of co-occurrences and false absences of interactions impact these metrics. %We find uncertainty influences both the average and the variance of network metrics, putting in perspective traditional network comparison studies where these metrics are assumed to be fixed entities.

%\section*{Case study: estimating the uncertainty of a plant-herbivore network}

    \subsection*{Description of the data}

      We use an extensively sampled \emph{Salix}-galler-parasitoid meta-network as a case study. This dataset consists of a single community type sampled across Europe; willow (\emph{Salix}) species, willow-galling sawflies (Hymenoptera, Tenthredinidae, Nematinae, Euurina), and their natural enemies (hymenopteran parasitoids and coleopteran, lepidopteran, dipteran, and hymenopteran inquilines). The data were collected over 29 years (1982-2010) at 374 unique locations across Europe ranging from Sicily to the Arctic. Including repeated visits to a number of sites, the data set contains 641 site-visits, each of which can be considered as a network in its own right. Here we consider each network as an independent sample from which to build the meta-network. The meta-network consists of 1,173 different interactions between 52 \emph{Salix} nodes, 92 herbivore nodes and 126 natural enemy nodes. Interactions were determined by dissecting and rearing gall inhabitants from 165,424 galls. The high spatiotemporal resolution of this network and the unusually high sampling effor implemented at the site level (641 site-visits, which can be considered as repeated samples of the network) makes this dataset particularly well suited for testing this method.

      Importantly, while large, this network illustrates the very real constraints on the sampling of ecological networks. As a baseline, we may compare the frequency of co-occurrence of taxa to reveal the challenge of having sufficiently large $n_{ij}$ (Fig.~\ref{upper_limits}). In the example data, most (3986 out of 4992) pairs of species are never found co-occurring, and for species found co-occurring the number of co-occurrences was generally low (Fig.~\ref{histograms}A). 
%DG: can you provide mean and variance in the text?
      Among the species which did sometimes interact when co-occurring, the incidence of interaction was actually low. 
%DG: in the following, would be useful to present both absolute and relative numbers, or just relative
      Only 141 pairs of species were observed interacting: of these, 26 were observed interacting only once and 48.2\% were observed interacting five or fewer times (Fig.~\ref{histograms}B). Thus, even in the most extensive data set that we could find, limited sampling is clearly a major source of uncertainty.

%DG: I really appreciate this figure, it makes the point so clear. 
%    \subsection*{Finding the maximum likelihood estimate}

@@

%DG: the different posterior connectance are the results of different samples of the posterior distribution, righ ?

>>>>

    The assumption of static communities conflicts with known variation in both  the nodes and links included in a network. Community composition and species abundances vary from site to site~\citep{Baiser2012} and over time within a site~\citep{Olesen2011a}. Likewise, interactions over space~\citep{Kitching1987,Baiser2012}, time~\citep{Kitching1987,Olesen2011a}, and between individuals of a given species~\citep{Pires2011a,Fodrie2015,Novak2015}. Moreover, variability in community composition and interactions may or may not be closely related. The removal of a species from a site will obviously also remove its interactions, but the co-occurrence of potentially interacting species do not guarantee they will interact at a given place and time. Interactions can be lost, or not detected, if participating species remain present but do not co-occur temporally, are too rare to detect each other~\citep{Tylianakis2010}, because of environmental contingencies \citep{Poisot2015}, or through changes to individual preferences~\citep{Fodrie2015}.

    Beyond ``true" variation in network structure, several researchers have pointed to the importance of sampling intensity for the assessment of network structure (e.g.,~\citealp{Martinez1999,Bluthgen2006,XX}). Accumulation of interactions with increasing sampling effort suggests that it is even more challenging to document interactions than species richness \citep{Poisot2012}. As a result, it was proposed that interactions should be described probabilistically and network metrics computed accordingly \citep{Poisot2016}. Despite this, the acknowledgment that insufficient sampling effort limits our ability to describe variation in ecological networks has not yet led to a quantitative methodology to deal with the uncertainty generated by spatiotemporal variation in ecological interactions and by sampling. Moreover, even in extremely well-sampled networks uneven sampling across species (or pairs of species) can lead to the assumption that some species do not interact because they co-occur rarely or have not yet been observed co-occuring (see Box 1). It is therefore likely that nearly all network studies are missing some interactions and could benefit from an approach that acknowledges this uncertainty.

@@

% DG : I wonder if a paragraph summarizing the extent of the problem with the salix data should not be provided here. Either within the text, or a box. But we need to convingly show how large is this problem even for such an extensive dataset
% AC : I like the box idea. Since we refer to the Salix data throughout the manuscript it makes sense to make it stand out.

    \indent As a case study, we use an extensively sampled \emph{Salix}-galler-parasitoid meta-network. This dataset consists of a single community type sampled across Europe; willow (\emph{Salix}) species, willow-galling sawflies, and their natural enemies. The data were collected over 29 years at 374 unique locations across Europe, with a total of 641 site visits. Each site visit can be considered as a network in its own right and an independent sample from which to build the meta-network. The meta-network consists of 1,173 different interactions between 52 \emph{Salix} nodes, 92 herbivore nodes, and 126 natural enemy nodes. The high spatiotemporal resolution of this network and the unusually high sampling effort implemented at the site level makes this dataset particularly well suited for illustrating the difficulties in completely sampling a network and testing Bayesian approaches to overcome these difficulties.

    \indent We may begin by comparing the frequency of of co-occurence of taxa to reveal the challenge of having sufficient sampling to be confident that an interaction does not occur (Fig.~\ref{upper_limits}). Most pairs of species (3986/4992 \emph{Salix}-galler pairs and  9794/12096 galler-natural enemy pairs) are never found co-occurring, and for species that did occur the number of co-occurrences was generally low (mean=4.24, var=36.3 for \emph{Salix}-galler pairs; mean=3.87, var=28.8 for galler-natural enemy pairs; Fig.~\ref{histograms}A-B). For those species which interacted when co-occurring, the incidence of interaction was also low (mean=12.0, var=155 for \emph{Salix}-galler pairs; mean=4.04, var=29.3 for galler-natural enemy pairs; Fig.~\ref{histograms}C-D). Only 2.82\% of \emph{Salix}-galler pairs and 7.76\% of galler-natural enemy pairs were observed interacting at all. Thus, even in the most extensive data set that we could find, limited sampling is clearly a major source of uncertainty.

@@

    The combination of these three sources of uncertainty together leads to many potential explanations for the observation of an absence of interaction, but only the situation where $L = 0$ (a true absence) is relevant for the investigation of networks. While the observation of an interaction is straightforward to interpret, absence of interactions must be decomposed into different quantities. It is particularly important to rule out the situations where $D=0 \cup X = 1 \cup L=1$, i.e. the interaction occurred at the location but was not observed, and $D=1 \cup X = 0 \cup L =1$, the interaction is feasible and would have been detected but did not occur at the local site. Ecologists seek to measure the occurrence of true absences, which is the joint event $L=0 \cup X=1 \cup D=1$), but in reality they measure the marginal probability $P(L) = k/n$ where $k$ is again the number of observed interactions and $n$ the number of observed co-occurrences.

@@

  An added complication is that not all sources of uncertainty are proportional to sample size. To record an interaction between A and B, we need to identify both partners correctly (a non-trivial problem in many food webs; e.g.~\citet{Kaartinen2011,Roslin2016}) and be able to resolve all interactions with a similar likelihood. For both molecular~\citep{} and rearing~\citep{} techniques, certain types of interactions may go unnoticed due to technical challenges. This can bias the set of recorded interactions.
  % DG : I propose to maintain part of this paragraph that was commented out, just to keep the transition. 
  The bottom line is that separating different sources of uncertainty is difficult indeed. As an alternative to abandoning empirical networks or continuing to ignore the uncertainty inherent in undetected observations, we propose that some insight regarding the detectability of interactions between species not found co-occurring may be gained from data on other species between which interactions are known to be feasible.

@@

          % [[Should we still have this? At the workshop we talked about how interaction probabilities depend on the degrees of both species and so I'm using predicted interaction probabilities (degreeA*degreeB) in the galler example. A writeup for that process follows the original text:]]

@@

          % [Tomas adds: This is a bit of a stump]
%DG: I tried to expand just a little, but I don't want to get lost in this paragraph as it would be a distraction. We should basically mention it as a possibility.

@@

    It is fairly straightforward to compute most of network metrics when the different $\lambda$ of the adjacency matrix are known, without variance~\citep{Poisot2016}. Several of these metrics derive directly from quantitative indices of network structure which are equivalent to $\lambda$; the remainder, originally defined for binary networks, can be adjusted to account for interaction probabilities between zero and one. It is not as easy, however, to understand how the uncertainty in these estimated interaction probabilities influences network metrics. Computation of these metrics involves non-linear functions and therefore, because of Jensen's inequality (which states that the average of non-linear function of a stochastic variable differs from the function of the average of that variable), any uncertainty in $\lambda$s could bias both the mean and variance of a network metric. For instance, assume that among all pairs of species with few observed co-occurrences there are many pairs which truly interact. If we begin with no observed links for these pairs, then correcting for these false absences will inflate the number of links and consequently influence metrics such as connectance, nestedness, and degree distribution. While we can, as described above, calculate these network metrics using fixed estimates for $\lambda$, no analytical solution currently exists for uncertain $\lambda$. Instead we must rely on computer simulations. In the following section, we will provide an empirical example, based on the data set published by~\citet{Kopelke2017} (see Box~\ref{box1}). We will show how the uncertainty arising from both absences of co-occurrences and false absences of interactions impact these metrics. %We find uncertainty influences both the average and the variance of network metrics, putting in perspective traditional network comparison studies where these metrics are assumed to be fixed entities.

@@

%DG: the different posterior connectance are the results of different samples of the posterior distribution, righ ? 
%Yes
